Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-03-26 analysis/paper/paper.Rmd
Removed sites with artefact count < 100 from the analysis

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

@@

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

>>>>

```{r libraries, eval = TRUE}
library(tidyverse)
library(FactoMineR)
library(factoextra)
```

@@

# Plots

```{r get-data, eval = TRUE}
sites <- read.csv(here::here('analysis/data/raw_data/Roalkvam_DialPast_March2021_data.csv'))

```

```{r wrangling}
sites <- sites %>% 
      # Find mean age for each site (to be used in the plot below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bp, reported_end_bp)),
         .keep = "unused") %>% 
  mutate(
    artefact_count = sum(c_across(core:flake_retouch_other))) %>% 
  filter(artefact_count > 100)

lithics <- sites %>%
  # Combine flint projectiles into a single variable
  mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch,
    macro_tool = axe + chisel, 
    macro_tool_other = axe_other + chisel_other + club_hatchet_other,
    small_tool_other = knife_other + scraper_other + drill_other + 
      microblade_retouch_other + blade_retouch_other + fragment_retouch_other +
      flake_retouch_other,
    .keep = "unused") %>%
  # Make site names the rownames
  column_to_rownames(var = "site") %>% 
  # Deselect a few of the columns not be used
  select(-one_of(c("n", "excavation_project", "excavated_m3")))
```

```{r cor, fig.cap = "Correspondence analysis using artefact count data.", fig.height = 5, fig.width = 10}
# Perform correspondence analysis on the data. 'mean_age' is meant for colouring
# so is excluded here
calithics <- CA(select(lithics, -mean_date), graph = FALSE)

# Find date to hold the central tendency date
centdate <- mean(lithics$mean_date) / 1000

# Call to plot
fviz_ca_row(calithics, col.row = lithics$mean_date / 1000, title = "",
                     repel = TRUE) +
  scale_color_gradient2(name = "Years cal. BP", low="blue", mid="yellow", 
                        high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

fviz_ca_col(calithics, col.col = "black", repel = TRUE, title = "")

```

Figure \@ref(fig:cor) displays a correspondence analysis using the lithic count data.

```{r pcadata}
# Tried to do this to avoid having to type the variable names into both mutate()
# and select(). The general syntax select(-(vector_of_variable_names)) worked,
# but not: mutate(new_varaible = sum(vector_of_variable_names) / artefact_count)
primary_lithics <- c("core", "core_fragment", "blade", "microblade", "chip",
                     "fragment", "flake", "flake_other", "fragment_other", 
                     "chip_other", "blade_other",  "microblade_other", 
                     "core_other", "core_fragment_other")

secondary_lithics <- c("axe_debitage", "axe", "chisel", "microlith", 
"tanged_point", "transverse_point", "single_edged_point", "hognipen_point",
"projectile_undefined", "microburin", "scraper", "knife", "burin", "drill", 
"blade_retouch", "microblade_retouch", "chip_retouch", "fragment_retouch",
"flake_retouch", "axe_other", "axe_debitage_other", "chisel_other",
"club_hatchet_other", "grinding_slab_stone", "projectile_point_other",
"knife_other", "scraper_other", "drill_other", "microblade_retouch_other", 
"blade_retouch_other", "fragment_retouch_other", "flake_retouch_other")

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
  # The reason for not using .keep below is for the option to retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake, flake_other, 
                     fragment_other, chip_other, blade_other,  microblade_other, 
                     core_other, core_fragment_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count) %>% 
  column_to_rownames("site")
```

```{r pca-analysis, fig.cap = "PCA.", fig.height = 7, fig.width = 10}
pcalithics <- PCA(select(lithics2, -mean_date), scale.unit = TRUE, graph = FALSE)
fviz_pca(pcalithics, col.ind = lithics2$mean_date / 1000, col.var = "black", 
         repel = TRUE) +  scale_color_gradient2(name = "Years cal. BP", 
         low="blue", mid="yellow", high="red", midpoint = centdate) +
  theme(legend.position = "bottom")

```

Figure \@ref(fig:pca-analysis) displays a principle components analysis using continuous measures as defined by @Bicho2020
